Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-04-14 Paper/Paper.tex
Updates from Overleaf

The interaction between technological innovation and labor markets has a direct impact across industries; job characteristics such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}). Labor market changes for one job type may have rippling effects on populations serviced by that industry. For example, physician labor markets directly affect patient accessibility and quality of care received. Physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations on physicians such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}). Thus, a shock to physician labor markets have potential implications for healthcare. I investigate physician labor market changes as a result of a major technology shift in the U.S. healthcare system, the implementation of electronic health records, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.

Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggestions for care. EHRs have have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2009 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), revealing the nationwide movement towards digitization.

Physicians, as primary users of this technology, play a meaningful role in whether potential benefits of EHRs are realized. With the rapid implementation of a complex technology, the practical tasks that take place in a physician's daily life changed just as rapidly. In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards other work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Treating EHR implementation in hospitals as exogenous to an individual physician, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, patients seen, and billing activity.

Using CMS Shared Patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians connected to hospitals. The data captures whether a physician in exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender. I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), (3) patients seen, and (4) billing activity. Common to each specification is the use of a binary treatment variable capturing a physician's EHR exposure in hospitals. In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR. A key assumption that underlies the main analysis is that exposure is exogenous to individual physicians. That is, the physician does not choose their own EHR exposure. While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making. For this reason, I include a second definition of EHR exposure that only considers exposure in hospitals where vertical integration between physician and hospital is low, as this indicates a much lower probability that EHR implementation is endogenous.

I do not observe granular details of what physicians do when they drop out of clinical settings, so I define retirement vaguely as the decision to cease clinical work altogether. Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation. Younger physicians were .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching. Whether physicians are young or old, these results suggest that access to care issues may have been exemplified as EHRs were implemented, and there was a surge of experienced physicians who left the labor force and were no longer available to influence younger physicians learning under them.

Most physicians did not choose to forego clinical work altogether. However, changing physical place of work might be an option if a physician is not satisfied with their current experience at a given hospital, whether it is because of burnout, loss of autonomy, or other factors. I investigate whether physicians switch from primarily hospital based work to office based settings, or more generally switch primary zip codes. I limit the sample to any physicians who did not retire and use the same event study framework to estimate the effect of EHR exposure on switching work settings. I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group. Further, when considering a more general outcome of whether a physician switches zip codes, I observe similar trends with even larger magnitudes. This implies that the technology imposed a large enough costs to induce switching and that physicians may be avoiding regulation that imposes loss of autonomy.

Finally, I investigate the effect of the technology implementation on work activity, measured by the number of patients seen, and overall billing activity by the physician. The subtle difference between the two measures represents an important nuance, that an increase in billing activity in itself could actually signify an increase in wastefulness as opposed to productivity if physicians are simply recording more in the EHR system. I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR. The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.

There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect. Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). More relevant to EHR effects on physicians, a number of studies consider the effect of EHRs on productivity in specific settings. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).

@@

\section{Institution}

EHRs have been an important feature of healthcare since the 1980s and 1990s. Early in its existence, the technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and/or scheduling. Prior to computerization, physicians were not able to interface directly with EHRs, and thus were not drastically affected by their implementation. The cost of utilizing an EHR was large and increasing in the amount of capabilities of the system, which led to the widespread view that EHRs were strictly complements to paper records, not a feasible substitute. As computers were made more portable, the usability of EHRs increased, creating what is known as the "physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates. Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry. Automation in data entry was non-existent, making it extremely time consuming for the user. Even when automation was developed for particular machines that performed monitoring, the hospital still held responsibility for the accuracy of information and therefore required physicians to manually check each data entry.

The HITECH Act was passed in 2009, setting aside government funds to subsidize hospitals who used EHRs according to certain guidelines. The guidelines include having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy. Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies. Hospitals began receiving payments for the first stage of meaningful use in 2011. This resulted in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). Figure \ref{fig:meanuse} shows a comparison of U.S. hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs. 2013.

While EHRs were widely utilized in hospitals by 2015, physicians still continue to report frustration. For every patient seen, a primary care physician spends an average of 16 minutes actively using an EHR, an amount of time that often exceeds face to face time with the patient (\cite{overhage2020physician}). The tasks that may take place during this time include viewing or inputting patient information, ordering services such as tests or prescriptions, navigating the system, or communication. Appendix () includes a computer screen image of what a physician typically sees when looking at an EHR.

@@

Using various linked datasets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time and other relevant characteristics. The different data sets used to construct the panel are described below.

@@

Medicare Data on Provider Practice and Specialty is a database of physicians which details claim counts and specialty from 2009-2017. The relevant information included is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room. I first limit the sample of physicians based on specialty type. Primary care physicians are exposed to the administrative burden of EHRs in hospitals more-so than specialists. Thus, I only include physicians whose primary specialty is reported as hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one). Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years, who likely do not have strong ties to the hospitals they have limited work in.

This data includes individual labor market information on physicians which I use to construct the dependent variables for my analysis. Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen and overall billing activity. These outcome variables and their construction are described in detail below.

\subsection{Shared Patient}

CMS Shared Patient data records annual information from 2009 to 2015 on the number of patients who bill two entities under Medicaid in the same day. For example, if a Medicaid patient is referred to a specialist by a primary care physician and the specialist is seen by the patient in the same day, then those two physicians have that shared patient in common. I utilize this data to explore physicians' EHR exposure based on the hospitals they share sufficient patients with. This measure of exposure defines various treatment variables for my analysis. I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS. I care specifically about primary care physicians who have a close working relationship with hospitals, who do rounds within at least one hospital. The types of physicians are already limited to primary care, but there still may be some office-based primary care physicians in the shared patient data. Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data. Specifically, a physician-hospital pair is dropped if their total number of same day billings with the hospital is less than 30 times the number of years the pair appears in the data. This is because it is important to avoid dropping physicians who leave the data for many years, as this may indicate retirement and skew the results of my analysis.

Using hospital NPI, I link the pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics. I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with. I create two measures of EHR exposure. The first is exposure to any EHR in any hospital, which attempts to capture the effect of first-time EHR use on physician labor markets. One concern with this approach is that a typical physician may be involved in the decision to implement an EHR and thus the treatment would be endogenous. While I cannot perfectly address this concern, I define an alternative treatment variable which theoretically will have less likelihood of this endogeneity problem. That is exposure to EHR at a "low integration" hospital (a hospital where physicians have a low measure of vertical integration with the hospital (\cite{madison2004hospital})). In low integrated settings, physicians do not have ownership or much control in the hospital setting. Results for this alternative defintion of exposure are presented in Appendix (). I then aggregate this data to the physician level. Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017. That is, the data contains no units which are "never-treated".

@@

\subsection{Outcome Variables}

The first dependent variable I consider is the decision for a physician to retire, or more broadly to stop practicing in a clinical setting altogether. Retirement is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. In words, retire is equal to one in the first year that a physician has no claim counts at any point in the future. I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement. Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results.

Next, I consider a physician switching place of work, such as switching to an office based setting or even selecting into different hospitals where preferences may be more aligned. I measure this in the data in three ways. I consider the fraction of patients a physician sees in an office based setting, which is directly available in the data. Further, I use this variable to construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office based setting. There is less variation in this data but allows insight to a different type of decision making where physicians may enter or leave office based setting based on EHR exposure. Lastly, I utilize the zip code information to define an indicator variable equal to one only if the physician's list of zip codes changes from one year to the next. While these variables do not give direct insight into whether physicians specifically move from one hospital to another, they do allow for exploration of whether settings are changing as a result of EHRs in general.

Finally, I consider outcome variables which explore productivity and billing activity once EHRs are implemented in hospitals. Productivity is measured in the number of patients a physician sees in a give year, where the billing activity is the total claim count for the physician. I distinguish between the two because it could be thd case that productivity does not actually increase, but EHRs could encourage excessive billing of things that would not have been billed for otherwise.

@@

Summary statistics for the entire sample are shown in Table \ref{tab:sumstats1}. Only 2\% of physicians retire over the course of the sample, which equates to approximately 750 physicians in the sample who retire at some point. A little over half of the sample works in an office in some capacity, with about a third of the number of patients being seen in an office setting. By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of being exposed. Physicians are typically exposed in any capacity about one fourth of the way through the sample and in low-integrated hospitals about halfway through the sample. The average physician is 48 years old, works with 1.5 hospitals and 1.3 systems.

I also include a table of means comparing physicians younger than 60, older than 60, and physicians who ever retire (who can fall in either age category) in Table \ref{tab:splitstats}. Senior physicians see more patients and work in more hospitals on average than younger physicians. These older physicians also see a larger fraction of patients in an office setting and are more likely to work in an office in general. Finally, physicians in the higher age bracket are less likely to switch zip codes. The age groups are exposed to EHRs in similar ways, but younger physicians are exposed to EHRs in low integrated setting slightly later. Retirement occurs just as often in the different age groups, indicating the possibility of physicians leaving a clinical setting for an administrative or other position.

In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time.. In 2009, 32\% of the sample of physicians were exposed to an EHR. That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period. Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed from 2015-2017. Exposure to EHRs only in low integrated hospitals follows a similar trend with lower exposure rates. When I utilize this treatment variable, I drop all physicians who are not exposed in a low-integration setting by 2015, so the sample of physicians changes according to this limitation. The black line represents the average fraction of a physician's hospitals which are utilizing EHRs over time. Not only are physicians gradually becoming exposed to EHRs, but their exposure is actually increasing over time.

@@

In this setting, treatment effects are likely heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues that arise in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all physicians who were treated, or first exposed to an EHR, in the same year. To estimate the heterogeneous treatment effects, I utilize the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).

There are several assumptions necessary to identify $ATT(g,t)$. First, I assume that treatment is not reversed once it occurs. That is, once a physician is exposed to an EHR, they cannot be un-exposed. This assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure. An EHR is a costly technology and requires a significant amount of work to implement. A hospital does not have incentive to un-implement an EHR once it is implemented. They may add features or switch vendors, but do not reverse EHR use. Further, since we think of treatment at the EHR level as exposure to the technology, this assumption is automatically satisfied when thought of as whether the physician has ever used an EHR. Once they are exposed to the EHR for the first time, that experience cannot be forgotten. Thus, physicians remain treated in all future years after exposure.

@@

As is usual in a difference in difference framework, I assume a version of parallel trends based on not-yet-treated units. I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups which were treated in later periods. This assumption would be violated under external conditions in certain years that may be correlated with labor markets, such as a major recession. The time period I study, 2009-2017, is reassuringly stable.

@@

A general concern in event study settings is whether pre-trends are driving the results. With each specification, I present the p-value for a Wald test for pre-trends. Visually, there are no clear indications of an increasing trend in retirement before exposure, and this is confirmed in the significantly large p-values.

@@

On average for the entire sample, claim counts are roughly 5.5 times the number of patients seen in a year. Shown in Figure \ref{fig:claim}, young physicians increase claim count by 37 in the year of exposure and 96 in the year after exposure, 3-4 times the increase in patient count. Thus, billing activity increases proportionally less than average. EHRs do not seem to be incentivizing physicians to bill more than they would have previously per patient.

>>>>

\section{Introduction}

How technological innovation affects labor markets is often controversial, in part because of the vast differences in technology and tasks present even within industries. For example, industrial robots decreased the share of hours worked in low-skilled workers from 1993-2007, but increased overall labor productivity  (\cite{graetz2018robots}), and even low-skilled workers in Norway who were not displaced by robots, but saw other workers displaced, experienced lower job satisfaction due to fear of displacement (\cite{schwabe2020automation}). Moreover, shocks to an existing labor market could have downstream effects on customers in that market. For example, technological innovation in health care may contribute to physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or technological change may encourage physician retirement, possibly limiting patient access to care and ultimately affecting patient health. In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S. health care system, the implementation of electronic health records.

Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggestions for care. EHRs have have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). Researchers and government officials expected the movement towards digitization to affect efficiency in health care substantially. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2009 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), revealing the nationwide movement towards digitization.

Physicians, as primary users of this technology, play an important role in whether potential benefits of EHRs are realized. With the rapid implementation of a complex technology, the practical tasks that take place in a physician's daily life changed just as rapidly. In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards other work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Using a difference-in-differences research design, I estimate the effects of hospital EHR implementation on physician labor market outcomes.

My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians who work exclusively in a hospital. The data captures whether a physician is exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender. I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and list of primary zip codes, (3) number of patients seen, and (4) billing activity. For all outcomes, the main independent variable of interest is a binary treatment variable capturing a physician's EHR exposure in hospitals. In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR.

Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation. Younger physicians (less than 60) were .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching. Whether physicians are young or old, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them. In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group. Further, when considering a more general outcome of whether a physician switches zip codes, I observe similar trends with even larger magnitudes. This implies that the technology imposed a large enough cost to induce switching and that physicians may be avoiding regulation that imposes loss of autonomy. Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR. The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.

A key assumption that underlies the main analysis is that exposure is exogenous to individual physicians. That is, the physician does not choose their own EHR exposure. While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making. For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a much lower probability that EHR implementation is endogenous. The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint decision making among physicians and hospitals regarding EHR adoptions.

My analysis contributes to the literature examining effects of health information technology on health care. There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect. Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). More relevant to EHR effects on physicians, a number of studies consider the effect of EHRs on productivity in specific settings. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).

@@

\section{Background on Electronic Health Records}

EHRs have been an important feature of healthcare since the 1980s and 1990s. Early in its existence, health care professionals perceived the technology as a complement to paper records, primarily deployed by large academic medical centers to improve efficiency in billing and/or scheduling. Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation. As computers were made more portable, the usability of EHRs increased, creating what is known as the "physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates. Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry. Automation in data entry was non-existent, making it extremely time consuming for the user. Even when automation was developed for particular machines that performed monitoring, the hospital still held responsibility for the accuracy of information and therefore required physicians to manually check each data entry (\cite{evans2016electronic}).

The HITECH Act was passed in 2009, designating \$27 billion in government subsidies to entities who used EHRs according to certain guidelines. The guidelines include having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy. The U.S. allocated subsidies according to stages of "meaningful" use: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, and stage 3 (2014) extended to data sharing between practices. This program was successful in  spurring EHR adoption, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). Figure \ref{fig:meanuse} shows a geographical comparison of U.S. hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs. 2013.

Figure \ref{fig:EPIC} shows a computer screen image of what a physician typically sees when viewing an EHR.

While most hospitals fully implemented EHRs by 2015, physicians still continue to report frustration. For every patient seen, a primary care physician spends an average of 16 minutes actively using an EHR, an amount of time that often exceeds face to face time with the patient (\cite{overhage2020physician}). The tasks that may take place during this time include viewing or inputting patient information, ordering services such as tests or prescriptions, navigating the system, or communication. Figure \ref{fig:EPIC} includes a computer screen image of what a physician typically sees when looking at an EHR.

@@

Using various linked data sets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, outcomes related to physician practice location and billable activity, and other relevant physician, hospital, and practice characteristics. I describe the different data sets used to construct the panel below.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room. I first limit the sample of physicians based on specialty type. My analysis relies on EHR implementation in hospitals. Thus, I only include physicians reported their primary specialty as hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one). Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.

This data includes information on physicians which I use to construct the dependent variables for my analysis. Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen, and overall billing activity. I describe these outcome variables and their construction in detail below in Section \ref{sec:outcome}.

\subsection{Shared Patients}

CMS Shared Patient data records annual information from 2009 to 2015 on the number of patients who bill two entities under Medicare. For example, if a primary care physician refers a patient to a specialist, then those two physicians have that particular shared patient in common. The number of shared patients are collected in 30, 90, or 180 day intervals. I focus on the number of shared patient billing two entities in the same day, within a 30 day interval. I utilize this data to explore physicians' EHR exposure based on the hospitals they share sufficient patients with. This measure of exposure defines various treatment variables for my analysis. I limit the entities by tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS. I am specifically interested in primary care physicians who have a close working relationship with hospitals, who do rounds within at least one hospital. The types of physicians are already limited to primary care, but there still may be office-based primary care physicians in the shared patient data. Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data. Specifically, a physician-hospital pair is dropped if their total number of same day billings with the hospital is less than 30 times the number of years the pair appears in the data. This is because it is important to avoid dropping physicians who leave the data for many years, as this may indicate retirement and skew the results of my analysis.

\subsection{AHA Survey Data}

Using hospital NPI, I link the physician-hospital pairs from the shared patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics. I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with. I create two measures of EHR exposure. The first is exposure to any EHR in any hospital, which attempts to capture the effect of first-time EHR use on physician labor markets. One concern with this approach is that a typical physician may be involved in the decision to implement an EHR and thus the treatment would be endogenous. While I cannot perfectly address this concern, I define an alternative treatment variable which theoretically will have less likelihood of this endogeneity problem. That is exposure to EHR at a "low integration" hospital (a hospital where physicians have a low measure of vertical integration with the hospital (\cite{madison2004hospital})). In low integrated settings, physicians do not have ownership or much control in the hospital setting. Results for this alternative defintion of exposure are presented in Appendix (). I then aggregate this data to the physician level. Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017. That is, the data contains no units which are "never-treated".

\subsection{Physician Compare}

@@

\subsection{Outcome Variables}\label{sec:outcome}

The first dependent variable I consider is the decision for a physician to retire, or more broadly to stop practicing in a clinical setting altogether. Retirement is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future. I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement. Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results.

Next, I consider a physician switching place of work, such as switching to an office based setting or even selecting into different hospitals where preferences may be more aligned. I measure this in the data in three ways. First, I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data. Second, I construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office-based setting. This binary measure of practice setting offers less variation but captures a more discrete changes in which physicians enter or leave the office-based setting entirely. Third, based on the physician's practice zip code, I form an indicator variable equal to one if the physician's list of zip codes changes from one year to the next. While no single variable directly measures physician movement across hospitals, these three outcomes collectively capture different ways in which physicians may change their practice setting over time.

Finally, I consider outcome variables which explore productivity and billing activity. Productivity is measured as the number of patients a physician sees in a given year, where the billing activity is the total claim count for the physician. I distinguish between the two because it could be the case that productivity does not actually increase, but EHRs could encourage more comprehensive billing.

@@

I show summary statistics for the entire sample in Table \ref{tab:sumstats1}. Only 2\% of physicians (approximately 750) retire over the course of the panel. A little over half of the sample works in an office in some capacity, with about a third of the number of patients being seen in an office setting. By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of being exposed. Physicians are typically exposed in any capacity about one fourth of the way through the sample and in low-integrated hospitals about halfway through the sample. The average physician is 48 years old, works with 1.5 hospitals and 1.3 systems.

I also include a table of means comparing physicians younger than 60, older than 60, and physicians who ever retire (who can fall in either age category) in Table \ref{tab:splitstats}. Senior physicians see more patients and work in more hospitals on average than younger physicians. These older physicians also see a larger fraction of patients in an office setting and are more likely to work in an office in general. Finally, physicians in the higher age bracket are less likely to switch zip codes. The age groups are exposed to EHRs in similar ways, but younger physicians are exposed to EHRs in low integrated setting slightly later.

In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time. In 2009, 32\% of the sample of physicians were exposed to an EHR. That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period. Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by 2015. Exposure to EHRs only in low integrated hospitals follows a similar trend with lower exposure rates. When I employ this treatment variable, I drop all physicians who are not exposed in a low-integration setting by 2015, so the sample of physicians changes according to this limitation. The black line represents the average fraction of a physician's hospitals which are utilizing EHRs over time. This shows that, along with becoming first exposed over the sample period, physicians are also seeing EHRs implemented in more of their own hospitals over time.

@@

In this setting, treatment effects are likely heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues that arise in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all physicians who were treated, or first exposed to an EHR, in the same year. To estimate the heterogeneous treatment effects, I employ the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).

There are several assumptions necessary to identify $ATT(g,t)$. First, I assume that treatment is not reversed once it occurs. That is, once a physician is exposed to an EHR, they cannot be un-exposed. This assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure. An EHR is a costly technology and requires a significant amount of collaboration to implement. A hospital does not have incentive to un-implement an EHR once it is implemented. They may add features or switch vendors, but do not reverse EHR use. Further, since we think of treatment at the EHR level as exposure to the technology, this assumption is automatically satisfied when thought of as whether the physician has ever used an EHR. Once they are exposed to the EHR for the first time, that experience cannot be forgotten. Thus, physicians remain treated in all future years after exposure.

@@

As is usual in a difference-in-differences framework, I assume a version of parallel trends based on not-yet-treated units. I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups which were treated in later periods. This assumption would be violated under external conditions in certain years that may be correlated with labor markets, such as a major recession. The time period I study, 2009-2017, is reassuringly stable.

@@

A general concern in event study settings is whether pre-trends are driving the results. With each specification, I present the p-value for a Wald test for pre-trends. Visually, there are no clear indications of an increasing trend in retirement before exposure, and this is confirmed statistically, where I fail to reject the null of parallel trends.

@@

On average for the entire sample, claim counts are roughly 5.5 times the number of patients seen in a year. Shown in Figure \ref{fig:claim}, young physicians increase claim count by 37 in the year of exposure and 96 in the year after exposure, 3-4 times the increase in patient count. Thus, billing activity increases proportionally less than average. EHRs do not seem to be facilitating more billable activity.
